---
title: "Basic R Commands"
author: "AnVILPublish"
date: "`r BiocStyle::doc_date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

This document demonstrates basic R commands and is intended as an example
input for the `as_notebook()` function in the **AnVILPublish** package. It
can be converted to a Jupyter notebook and used interactively in an AnVIL
workspace.

# Arithmetic and Assignment

R can be used as a calculator. Results are assigned to variables with `<-`.

```{r arithmetic}
x <- 5 + 3
y <- 10 / 2
z <- x * y
z
```

# Vectors

Vectors are the basic data structure in R, created with `c()`.

```{r vectors}
nums <- c(1, 4, 9, 16, 25)
nums

# Vectorized operations
sqrt(nums)
nums * 2
```

# Summary Statistics

```{r summary}
values <- c(3.1, 7.4, 2.9, 8.6, 5.0, 4.2, 6.8)

mean(values)
median(values)
sd(values)
range(values)
summary(values)
```

# Data Frames

Data frames store tabular data. Here we create a small one from scratch.

```{r data-frame}
df <- data.frame(
    name  = c("Alice", "Bob", "Carol", "Dave"),
    score = c(88, 73, 95, 61),
    grade = c("B", "C", "A", "D")
)
df
```

Access columns with `$` and filter rows with logical indexing.

```{r data-frame-access}
# Column access
df$score

# Filter rows where score is above 75
df[df$score > 75, ]
```

# Basic Visualization

R's built-in `plot()` function produces quick graphics.

```{r plot}
x_vals <- seq(0, 2 * pi, length.out = 100)
y_vals <- sin(x_vals)

plot(
    x_vals, y_vals,
    type = "l",
    xlab = "x", ylab = "sin(x)",
    main = "Sine Wave"
)
```

A simple bar chart of the scores from the data frame above:

```{r barplot}
barplot(
    df$score,
    names.arg = df$name,
    ylab = "Score",
    main = "Student Scores",
    col  = "steelblue"
)
```

# String Manipulation

```{r strings}
greeting <- "Hello, AnVIL!"
nchar(greeting)
toupper(greeting)
gsub("AnVIL", "World", greeting)
```

# Control Flow

```{r control-flow}
for (i in 1:5) {
    if (i %% 2 == 0) {
        message(i, " is even")
    } else {
        message(i, " is odd")
    }
}
```

# Functions

User-defined functions encapsulate reusable logic.

```{r functions}
square_root_sum <- function(a, b) {
    sqrt(a) + sqrt(b)
}

square_root_sum(9, 16)
square_root_sum(2, 8)
```

# Session Information

```{r session-info}
sessionInfo()
```
